---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Let´s set the global options. We set the option cache = TRUE to avoid compiling all the code chunks again and again. The plots are goint to
be placed in the Figure folder.
```{r setoptions, echo = TRUE}
library(knitr)
opts_chunk$set(echo = TRUE, cache = TRUE, fig.path = "Figure/")
```
Set the working directory and read the .csv file
```{r read_file}
setwd("~/3. Personal/JHU - Data Science/5. Reproducible Research/Assignment week 2/RepData_PeerAssessment1")
activity <- read.csv("activity.csv")
```
        
## What is mean total number of steps taken per day?

First, we ignore the missing values
```{r missing_values}
c_activity <- activity[complete.cases(activity),]
```
Calculate the total number of steps per day
```{r number_steps}
library(dplyr)
t_steps <- c_activity %>% group_by(date) %>% summarize(total = sum(steps))
```
Make a histogram of the total number of steps taken each day
```{r histogram}
hist(t_steps$total, col = "red", xlab = "Steps", ylab = "Number of days", main = "Histogram of the total number of steps/day")

```

Calculate and report the mean and median of the total number of steps taken per day
```{r mean_and_median}
summary(t_steps$total)
```
The mean of the total number of steps taken per day is **10.766**. And the median is **10.765**.


## What is the average daily activity pattern?

First, we group by interval and calculate the total number of steps each 5 min. interval 
```{r average_daily_activity}
library(dplyr)
interval <- c_activity %>% group_by(interval) %>% summarize(average = mean(steps))
```
Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r time_series_plot}

plot(x = interval$interval, y = interval$average/mean(c_activity$steps), xlab = "Interval", ylab = "Average of steps taken", 
     type = "l", main = "Average steps taken per interval")

max_steps <-  max(interval$average)
max_inter <- as.numeric(interval[interval$average == max_steps, 1])
max_inter
```

On average across all the days in the dataset, the 5-minute interval which contains the maximum number of steps is **`r max_inter`**

## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r number_of_NAs}
number_NA <- nrow(activity) - nrow(c_activity)
number_NA
```
The total number of missing values is **`r number_NA`**

Fill in all of the missing values in the dataset with the mean for that 5-minute interval.
```{r filling_NAs}
fill_activity <- activity
for (i in 1:nrow(activity))
{
        if (is.na(activity[i, 1])) 
        {
                fill_activity[i,1] <- interval[interval$interval == activity[i,3], 2]
        }
}

```

Make a histogram of the total number of steps taken each day
```{r histogram_NA}
library(dplyr)
t_steps_NA <- fill_activity %>% group_by(date) %>% summarize(total = sum(steps))
hist(t_steps_NA$total, col = "red", xlab = "Steps", ylab = "Number of days", main = "Histogram of the total number of steps/day")

```


Calculate and report the mean and median total number of steps taken per day. 
```{r mean_and_median_NA}
summary(t_steps_NA$total)
```
The mean and the median total number of steps per day is **10.766**. When imputing missing data, the median increases a little bit.



## Are there differences in activity patterns between weekdays and weekends?

### Create a new factor variable in the dataset with two levels – “weekday” and “weekend” 

Transform the date colunm in date format
```{r Convert_dates}
fill_activity$date <- as.Date(fill_activity$date)
```

Identify which date is weekend or weekday (note that in my computer R is set up in Spanish)
```{r weekday}
library(lubridate)
library(dplyr)

for (i in 1:nrow(fill_activity))
{
      if (weekdays(fill_activity$date[i]) == "sábado" || weekdays(fill_activity$date[i]) == "domingo")
      {
          fill_activity$week[i] <- "weekend"
      }
      else {fill_activity$week[i] <- "weekday"}   
}
```

Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r plot_weekend}
weekdays <- fill_activity[fill_activity$week == "weekday", ]
weekends <- fill_activity[fill_activity$week == "weekend", ]

library(dplyr)
avg_weekdays <- weekdays %>% group_by(interval) %>% summarize(average = mean(steps))
avg_weekends <- weekends %>% group_by(interval) %>% summarize(average = mean(steps))

par(mfrow = c(2, 1), mar = c(4, 4, 1, 1), oma = c(0, 0, 0, 0))

plot(x = avg_weekdays$interval, y = avg_weekdays$average/mean(weekdays$steps), xlab = "", ylab = "Number of steps", 
     type = "l", main = "weekdays", ylim = range(c(0,8)))

plot(x = avg_weekends$interval, y = avg_weekends$average/mean(weekends$steps), xlab = "Interval", ylab = "Number of steps", 
     type = "l", main = "weekends", ylim = range(c(0,8)))


```




















